---
title: "Lab 4 code"
output: html_document
date: "2025-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in your data. Use the extra dataset for any questions you're not able to answer with your own data.
Edit the code below to read in your data
```{r reading_in_data}

control_data <- read.csv("/Users/sarahlapidus/Documents/Malawi training docs/Data sets/Data for github/simulated_control_long_training_data.csv")
sample_data <- read.csv("/Users/sarahlapidus/Documents/Malawi training docs/Data sets/simulated_sample_wide_training_data.csv")

#this converts sample_data from a wide to a long dataframe
sample_long <- reshape(
  sample_data,
  varying = setdiff(names(sample_data), c("id", "age", "sex")),
  v.names = "mfi",
  timevar = "antigen",
  times = setdiff(names(sample_data), c("id", "age", "sex")),
  idvar = "id",
  direction = "long"
)
rownames(sample_long) <- NULL


```

1. What positive and/or negative controls do you have available? What was the source of these controls?

2. Using unexposed controls, adjust the code to do the following: 
  a. Determine whether your negative controls are normally distributed on the natural or log scale. 
  
```{r histogram_neg_controls}

    #restrict to negative controls
    control_negs <- control_data[control_data$pos_neg=="negative",]

    #check distributions of data - untransformed
    neg_controls_natural_scale <- ggplot(control_negs, aes(x = mfi)) +
      geom_histogram(bins = 10, color = "black", fill = "blue") +
      facet_wrap(~ antigen, scales = "free", ncol = 4) +  # <- Set 5 columns per row
      labs(
        title = "Sample Distribution (Natural Scale)",
        x = "Antigen MFI",
        y = "Number of samples"
      ) +
      theme_minimal() +
      theme(
        strip.text = element_text(size = 10),  # Smaller font for facet labels
        axis.text = element_text(size = 10),  # Smaller font for axis text
        plot.title = element_text(size = 10)  # Smaller font for the plot title
      )
    neg_controls_natural_scale
    
    neg_controls_log_scale <- ggplot(control_negs, aes(x = log(mfi))) +
      geom_histogram(bins = 10, color = "black", fill = "blue") +
      facet_wrap(~ antigen, scales = "free", ncol = 4) +  # <- Set 5 columns per row
      labs(
        title = "Sample Distribution (Log Scale)",
        x = "Log Antigen MFI",
        y = "Number of samples"
      ) +
      theme_minimal() +
      theme(
        strip.text = element_text(size = 10),  # Smaller font for facet labels
        axis.text = element_text(size = 10),  # Smaller font for axis text
        plot.title = element_text(size = 10)  # Smaller font for the plot title
      )
    neg_controls_log_scale
```
  
  
  b. Based on the histograms above, chose an antigen with relatively normally distributed negative controls (untransformed or on log scale) for which it could be appropriate to use the mean + SD cutoff method.
  Calculate a cutoff value as mean + 2 or 3 standard deviations. Use a natural or log transformation as appropriate. Adjust the code below. 
  
```{r mean_sd_cutoffs}

  #for ZIKA on log scale. Adjust for your own file
  ZIKA_control_negs <- control_negs[control_negs$antigen == "ZIKA","mfi"] #creates a vector of mfi values for ZIKA
  ZIKA_control_negs_log <- log(ZIKA_control_negs)

  #calculates cutoffs as mean +2 or 3 SDs. Since we did this on the log scale, we exponentiate results for cutoff on untransformed scale.
  ZIKA_cutoff_2sd <- exp(mean(ZIKA_control_negs_log, na.rm = TRUE) + 2 * sd(ZIKA_control_negs_log, na.rm = TRUE))
  ZIKA_cutoff_3sd <- exp(mean(ZIKA_control_negs_log, na.rm = TRUE) + 3 * sd(ZIKA_control_negs_log, na.rm = TRUE))
  print(ZIKA_cutoff_2sd)
  print(ZIKA_cutoff_3sd)


```

  c. How are your negative controls similar/different to negatives in your population?
  
  d. What is the seropositivity you calculated using negative controls as a cutoff?
  
```{r seropositivity_mean_sd}

  ZIKA_control_negs <- control_negs[control_negs$antigen == "ZIKA","mfi"] #creates a vector of mfi values for ZIKA

  #this code applies the cutoffs we found above to define people as seropositive or seronegative
  #adjust with the names of your cutoffs
  seropositivity_2sd <- ifelse(sample_data$ZIKA > ZIKA_cutoff_2sd, 1, 0) #1 is seropositive and 0 is seronegative 
  seropositivity_3sd <- ifelse(sample_data$ZIKA > ZIKA_cutoff_3sd, 1, 0)
  
  #gives numbers of seropositive and negative 
  table(seropositivity_2sd,useNA="always")
  table(seropositivity_3sd,useNA="always")  
  
  #gives percent of people seropositive and negative
  round(prop.table(table(seropositivity_2sd,useNA="always")),3)*100
  round(prop.table(table(seropositivity_3sd,useNA="always")),3)*100

```
  
3. Use the ROC method to adjust the code to do the following: 
  a. Choose an MFI value and calculate sensitivity and specificity using positive and negative controls for an antigen. 
  
```{r sensitivity_specificity}

cutoff <- 600 #adjust as you choose 

#restricting just to controls for the antigen we're looking at. Adjust for your own antigen
control_data_Zika <- control_data[which(control_data$antigen == "ZIKA"),1:3]

# Create Seropos variable based on MFI cutoff
seropositivity <- ifelse(control_data_Zika$mfi > cutoff, 1, 0)  

#Since we have positive and negative controls, we know which samples are true positive and negatives (from pos_neg variable), compared to the samples we've assigned as positive or negative (based on seropositivity)
table(control_data_Zika$pos_neg, seropositivity)

#Calculate True Positives, True Negatives,  False Positives, and False Negatives
#in your dataset, check if pos_neg values are coded as positive/negative, pos/neg, 1/0 etc, and change accordingly
TP <- sum(seropositivity == 1 & control_data_Zika$pos_neg == "positive")  # True Positives
TN <- sum(seropositivity == 0 & control_data_Zika$pos_neg == "negative")  # True Negatives
FP <- sum(seropositivity == 1 & control_data_Zika$pos_neg == "negative")  # False Positives
FN <- sum(seropositivity == 0 & control_data_Zika$pos_neg == "positive")  # False Negatives

# Calculate Sensitivity and Specificity
Sensitivity <- TP / (TP + FN)
Specificity <- TN / (TN + FP)

# Print results
cat("True Positives:", TP, "\n")
cat("True Negatives:", TN, "\n")
cat("False Positives:", FP, "\n")
cat("False Negatives:", FN, "\n")
cat("Sensitivity:", round(Sensitivity, 3), "\n")
cat("Specificity:", round(Specificity, 3), "\n")


```

  b. Plot an ROC curve for this antigen.


```{r generate_roc_curve}

# Generate the ROC curve
#compute_roc is a function in the source file. Put in a vector of which controls are positive and negative, and a vector of mfi for each control, and it returns a dataframe with False Positive Rate (FPR) and True Positive Rate (TPR) at each cutoff
source("/Users/sarahlapidus/Documents/Malawi training docs/Source file V1.R")
roc_df <- compute_roc(control_data_Zika$pos_neg, control_data_Zika$mfi)

#we can then plot the results of this function using ggplot to make an ROC curve
ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = "ROC Curve",
    x = "False Positive Rate (FPR)",
    y = "True Positive Rate (TPR)"
  ) +
  theme_minimal()
```

  c. How will you determine the best place on the ROC curve to establish a cutoff? What does this mean for how you weigh possible misclassification?
  
  d. Using the ROC curve and the roc_df you generated above, what cutoff value did you use? You want to find a TPR closest to 1 and an FPR closest to 0.  Calculate seropositivity using this cutoff.
  
```{r apply_roc_cutoff}

roc_cutoff <- 550 ##put the cutoff you chose here
seropositivity <- ifelse(sample_data$ZIKA > roc_cutoff, 1, 0) 
 
#number of seropositive people 
table(seropositivity,useNA="always")  
  
#percent of people seropositive and negative
round(prop.table(table(seropositivity,useNA="always")),3)*100
  
```


e. Adjust the code below to calculate Youden's J statistic. This is the cutoff that maximizes the value of: Sensitivity + Specificity – 1

```{r Youden's_J}
# Calculate Youden's J statistic
roc_df$YoudenJ <- roc_df$TPR - roc_df$FPR

# Find the row with the maximum Youden's J
best_index <- which.max(roc_df$YoudenJ)
youden_cutoff <- roc_df$Threshold[best_index]

# Optionally print the cutoff and corresponding TPR/FPR
cat("Youden cutoff:", youden_cutoff, "\n")
cat("TPR at cutoff:", roc_df$TPR[best_index], "\n")
cat("FPR at cutoff:", roc_df$FPR[best_index], "\n")

```

4. Optional: Plot your sample values and consider if they have a good distribution to fit a mixture model.

```{r sample_histogram}

  faceted_natural_scale <- ggplot(sample_long, aes(x = mfi)) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  facet_wrap(~ antigen, scales = "free", ncol = 5) +  # <- Set 5 columns per row
  labs(
    title = "",
    x = "Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_natural_scale

#log scale
faceted_log_scale <- ggplot(sample_long, aes(x = log(mfi))) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  facet_wrap(~ antigen, scales = "free", ncol = 5) +  # <- Set 5 columns per row
  labs(
    title = "",
    x = "Log Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_log_scale

```

5. Compare results of all the cutoff methods you have applied. 

  a. For one antigen, adjust the code to plot the sample results with the different cutoffs indicated. Do this on the natural and log scale
  
```{r natural scale}

#cutoffs are: 
# ZIKA_cutoff_2sd
# ZIKA_cutoff_3sd
# roc_cutoff

faceted_natural_scale <- ggplot(sample_data, aes(x = WMEV)) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  geom_vline(xintercept = ZIKA_cutoff_2sd, linetype = "longdash", color = "red", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_2sd
  geom_vline(xintercept = ZIKA_cutoff_3sd, linetype = "longdash", color = "purple", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_3sd
  geom_vline(xintercept = roc_cutoff, linetype = "longdash", color = "darkgreen", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  geom_vline(xintercept = youden_cutoff, linetype = "longdash", color = "orange", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  labs(
    title = "Put your title here",
    x = "Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_natural_scale


```
  
```{r log_scale}
#log scale
faceted_log_scale <- ggplot(sample_data, aes(x = log(ZIKA))) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  geom_vline(xintercept = log(ZIKA_cutoff_2sd), linetype = "longdash", color = "red", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_2sd
  geom_vline(xintercept = log(ZIKA_cutoff_3sd), linetype = "longdash", color = "purple", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_3sd
  geom_vline(xintercept = log(roc_cutoff), linetype = "longdash", color = "darkgreen", size = 0.8) +  # Add vertical dashed line at roc_cutoff
    geom_vline(xintercept = log(youden_cutoff), linetype = "longdash", color = "orange", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  labs(
    title = "Put your title here",
    x = "Log Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10), 
    axis.text = element_text(size = 10),  
    plot.title = element_text(size = 10) 
  )
faceted_log_scale
```  


  b. How similar / different are the cutoff methods?


6. Which method do you think you will use going forward in your study (if you will use a cutoff)? Why? 
