---
title: "Lab 4 code"
output: html_document
date: "2025-05-22"
---

Lab 4 objective: Apply different cutoff methods to calculate seroprevalence and compare the results. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r source_and_working_directory}
source("~/seroanalytics_workshop/Source file V1.R")
```

Read in your data. Use the extra dataset for any questions you're not able to answer with your own data.
Edit the code below to read in your data
```{r reading_in_data}

standards_data <- read.csv("data/simulated_standard_curve_training_data.csv")
control_data <- read.csv("data/simulated_control_long_training_data.csv")
sample_data <- read.csv("data/simulated_sample_wide_training_data.csv")

#this converts sample_data from a wide to a long dataframe. edit the column names to your data.
sample_long <- reshape(
  sample_data,
  varying = setdiff(names(sample_data), c("id", "age", "sex")),
  v.names = "mfi",
  timevar = "antigen",
  times = setdiff(names(sample_data), c("id", "age", "sex")),
  idvar = "id",
  direction = "long"
)
rownames(sample_long) <- NULL


```

1. What positive and/or negative controls do you have available for which antigens? What was the source of these controls?

2. Using unexposed controls, adjust the code to do the following: 
  a. Determine whether your negative controls are normally distributed on the natural or log scale. 
  
```{r histogram_neg_controls}

    #restrict to negative controls
    control_negs <- control_data[control_data$pos_neg=="negative",] #edit this line if needed

    #check distributions of data - untransformed
    neg_controls_natural_scale <- ggplot(control_negs, aes(x = mfi)) +
      geom_histogram(bins = 10, color = "black", fill = "blue") +
      facet_wrap(~ antigen, scales = "free", ncol = 4) +  # <- Set 5 columns per row
      labs(
        title = "Sample Distribution (Natural Scale)",
        x = "Antigen MFI",
        y = "Number of samples"
      ) +
      theme_minimal() +
      theme(
        strip.text = element_text(size = 10),  # Smaller font for facet labels
        axis.text = element_text(size = 10),  # Smaller font for axis text
        plot.title = element_text(size = 10)  # Smaller font for the plot title
      )
    neg_controls_natural_scale
    
    neg_controls_log_scale <- ggplot(control_negs, aes(x = log(mfi))) +
      geom_histogram(bins = 10, color = "black", fill = "blue") +
      facet_wrap(~ antigen, scales = "free", ncol = 4) +  # <- Set 5 columns per row
      labs(
        title = "Sample Distribution (Log Scale)",
        x = "Log Antigen MFI",
        y = "Number of samples"
      ) +
      theme_minimal() +
      theme(
        strip.text = element_text(size = 10),  # Smaller font for facet labels
        axis.text = element_text(size = 10),  # Smaller font for axis text
        plot.title = element_text(size = 10)  # Smaller font for the plot title
      )
    neg_controls_log_scale
```
  
  b. Based on the histograms above, chose an antigen for which it could be appropriate to use the mean + SD cutoff method.
  Calculate a cutoff value as mean + 2 or 3 standard deviations. Decide whether you want to use untransformed or log transformed data. Adjust the code below. 
  
```{r mean_sd_cutoffs}

  #for ZIKA on log scale. Adjust for your own file
  ZIKA_control_negs <- control_negs[control_negs$antigen == "ZIKA","mfi"] #creates a vector of mfi values for ZIKA
  ZIKA_control_negs_log <- log(ZIKA_control_negs)

  #calculates cutoffs as mean +2 or 3 SDs. Since we did this on the log scale, we exponentiate results for cutoff on untransformed scale.
  ZIKA_cutoff_2sd <- exp(mean(ZIKA_control_negs_log, na.rm = TRUE) + 2 * sd(ZIKA_control_negs_log, na.rm = TRUE))
  ZIKA_cutoff_3sd <- exp(mean(ZIKA_control_negs_log, na.rm = TRUE) + 3 * sd(ZIKA_control_negs_log, na.rm = TRUE))
  print(ZIKA_cutoff_2sd)
  print(ZIKA_cutoff_3sd)

```

  c. How are your negative controls similar/different to negatives in your population?
  
  d. What is the seropositivity you calculated using negative controls as a cutoff?
  
```{r seropositivity_mean_sd}

  ZIKA_control_negs <- control_negs[control_negs$antigen == "ZIKA","mfi"] #creates a vector of mfi values for ZIKA

  #this code applies the cutoffs we found above to define people as seropositive or seronegative
  #adjust with the names of your cutoffs
  seropositivity_2sd <- ifelse(sample_data$ZIKA > ZIKA_cutoff_2sd, 1, 0) #1 is seropositive and 0 is seronegative 
  seropositivity_3sd <- ifelse(sample_data$ZIKA > ZIKA_cutoff_3sd, 1, 0)
  
  #gives numbers of seropositive and negative 
  cat("Table of number seronegative and seronegative\n")
  table(seropositivity_2sd,useNA="always")
  table(seropositivity_3sd,useNA="always")  
  
  #gives percent of people seropositive and negative
  cat("Table of percent seronegative and seronegative\n")
  round(prop.table(table(seropositivity_2sd,useNA="always")),3)*100
  round(prop.table(table(seropositivity_3sd,useNA="always")),3)*100

```

  e. Calculate confidence intervals for this seroprevalence.
  
```{r mean_sd_confidence_interval}

# for Mean + 2SD (you can adjust code for mean + 3SD)
  mean_2sd_table <- table(seropositivity_2sd,useNA="always")
  mean_2sd_table <- as.data.frame(mean_2sd_table)
  
x <- mean_2sd_table$Freq[2]
n <- sum(mean_2sd_table$Freq)
conf <- 0.95

# Exact interval
mean_sd_ci <- binom.exact(x, n, conf.level = conf) #epitools function
cat("CI lower", round(mean_sd_ci$lower,4)*100, "%, CI upper", round(mean_sd_ci$upper,4)*100, "%")

```

3. Sensitivity and specificity

  a. Select a reasonable cutoff for an antigen with positive and negative controls. For this cutoff method, find the number of true postives, true negatives, false positives, and false negatives
  
```{r 2x2 table}

cutoff <- 600 #adjust as you choose 

#restricting just to controls for the antigen we're looking at. Adjust for your own antigen
control_data_Zika <- control_data[which(control_data$antigen == "ZIKA"),1:3]

# Create Seropos variable based on MFI cutoff
seropositivity <- ifelse(control_data_Zika$mfi > cutoff, 1, 0)  

#Since we have positive and negative controls, we know which samples are true positive and negatives (from pos_neg variable), compared to the samples we've assigned as positive or negative (based on seropositivity)
cat("2X2 table showing the positive and negative controls by serostatus", "\n")
table(control_data_Zika$pos_neg, seropositivity)

#Calculate True Positives, True Negatives,  False Positives, and False Negatives
#in your dataset, check if pos_neg values are coded as positive/negative, pos/neg, 1/0 etc, and change accordingly
TP <- sum(seropositivity == 1 & control_data_Zika$pos_neg == "positive")  # True Positives
TN <- sum(seropositivity == 0 & control_data_Zika$pos_neg == "negative")  # True Negatives
FP <- sum(seropositivity == 1 & control_data_Zika$pos_neg == "negative")  # False Positives
FN <- sum(seropositivity == 0 & control_data_Zika$pos_neg == "positive")  # False Negatives

cat("\n","Number of True positives:", TP, "\n")
cat("Number of True negative:", TN, "\n")
cat("Number of False positives:", FP, "\n")
cat("Number of False negatives:", FN, "\n")

```

  b. calculate sensitivity and specificity using positive and negative controls for this antigen. 
  
```{r sensitivity_and_specificity}

# Calculate Sensitivity and Specificity
Sensitivity <- TP / (TP + FN)
Specificity <- TN / (TN + FP)

# Print results
cat("True Positives:", TP, "\n")
cat("True Negatives:", TN, "\n")
cat("False Positives:", FP, "\n")
cat("False Negatives:", FN, "\n")
cat("Sensitivity:", 100*round(Sensitivity, 3), "%\n")
cat("Specificity:", 100*round(Specificity, 3), "%\n")
```

  c. Plot an ROC curve for this antigen. Edit code to use your own antigen
    replace control_data_Zika$mfi with your vector of control mfi values
    replace control_data_Zika$pos_neg with your vector that shows which are positive and negative controls

```{r generate_roc_curve}

# Step 1: Create thresholds
thresholds <- sort(unique(control_data_Zika$mfi), decreasing = TRUE)

# Step 2: Compute TPR (sensitivity) and FPR (1 - specificity) for each threshold
roc_df <- data.frame(threshold = thresholds, TPR = NA, FPR = NA)

for (i in seq_along(thresholds)) {
  t <- thresholds[i]
  predicted <- ifelse(control_data_Zika$mfi >= t, 1, 0)
  
  TP <- sum(predicted == 1 & control_data_Zika$pos_neg == "positive")  # True Positives
  TN <- sum(predicted == 0 & control_data_Zika$pos_neg == "negative")  # True Negatives
  FP <- sum(predicted == 1 & control_data_Zika$pos_neg == "negative")  # False Positives
  FN <- sum(predicted == 0 & control_data_Zika$pos_neg == "positive")  # False Negatives
    
  TPR <- TP / (TP + FN)
  FPR <- FP / (FP + TN)
  
  roc_df$TPR[i] <- TPR
  roc_df$FPR[i] <- FPR
}

# Step 3: Calculate AUC using the trapezoidal rule
roc_df <- roc_df[order(roc_df$FPR), ]
auc <- sum(diff(roc_df$FPR) * (head(roc_df$TPR, -1) + tail(roc_df$TPR, -1)) / 2)

# Step 4: Plot ROC curve using ggplot2
ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(title = paste0("ROC Curve (AUC = ", round(auc, 3), ")"),
       x = "False Positive Rate (1 - Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_minimal()
```

  d. How will you determine the best place on the ROC curve to establish a cutoff? What does this mean for how you weigh possible misclassification (ie do you want to prioritize minimizing false negatives, false positives, or both?
  
  e. Using the ROC curve and the roc_df you generated above, what cutoff value did you use? You want to find a TPR closest to 1 and an FPR closest to 0.  Calculate seropositivity using this cutoff.
  
```{r apply_roc_cutoff}

roc_cutoff <- 550 ##put the cutoff you chose here
seropositivity <- ifelse(sample_data$ZIKA > roc_cutoff, 1, 0) 
 
#number of seropositive people 
table(seropositivity,useNA="always")  
  
#percent of people seropositive and negative
round(prop.table(table(seropositivity,useNA="always")),3)*100
  
```

  f. Calculate confidence intervals for this seroprevalence.
  
```{r roc_confidence_interval}

# for Mean + 2SD (you can adjust code for mean + 3SD)
  roc_table <- table(seropositivity,useNA="always")  
  roc_table <- as.data.frame(roc_table)
  
x <- roc_table$Freq[2]
n <- sum(roc_table$Freq)
conf <- 0.95

# Exact interval
ci_roc <- binom.exact(x, n, conf.level = conf) #epitools function
cat("CI lower", round(ci_roc$lower,4)*100, "%, CI upper", round(ci_roc$upper,4)*100, "%")

```

  g. Adjust the code below to calculate Youden's J statistic. This is the cutoff that maximizes the value of: Sensitivity + Specificity – 1
  This is the cutoff you would choose if you want to minimize both false positives and false negatives

```{r Youden's_J}
# Calculate Youden's J statistic
roc_df$YoudenJ <- roc_df$TPR - roc_df$FPR

# Find the row with the maximum Youden's J
best_index <- which.max(roc_df$YoudenJ)
youden_cutoff <- roc_df$threshold[best_index]

# Optionally print the cutoff and corresponding TPR/FPR
cat("Youden cutoff:", youden_cutoff, "\n")
cat("TPR at cutoff:", roc_df$TPR[best_index], "\n")
cat("FPR at cutoff:", roc_df$FPR[best_index], "\n")

```

4. Compare results of all the cutoff methods you have applied. 

  a. For one antigen, adjust the code to plot the sample results with the different cutoffs indicated. Do this on the natural and log scale
  
```{r natural scale}

#cutoffs are: 
# ZIKA_cutoff_2sd
# ZIKA_cutoff_3sd
# roc_cutoff

faceted_natural_scale <- ggplot(sample_data, aes(x = WMEV)) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  geom_vline(xintercept = ZIKA_cutoff_2sd, linetype = "longdash", color = "red", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_2sd
  geom_vline(xintercept = ZIKA_cutoff_3sd, linetype = "longdash", color = "purple", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_3sd
  geom_vline(xintercept = roc_cutoff, linetype = "longdash", color = "darkgreen", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  geom_vline(xintercept = youden_cutoff, linetype = "longdash", color = "orange", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  labs(
    title = "Put your title here",
    x = "Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_natural_scale


```
  
```{r log_scale}
#log scale
faceted_log_scale <- ggplot(sample_data, aes(x = log(ZIKA))) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  geom_vline(xintercept = log(ZIKA_cutoff_2sd), linetype = "longdash", color = "red", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_2sd
  geom_vline(xintercept = log(ZIKA_cutoff_3sd), linetype = "longdash", color = "purple", size = 0.8) +  # Add vertical dashed line at ZIKA_cutoff_3sd
  geom_vline(xintercept = log(roc_cutoff), linetype = "longdash", color = "darkgreen", size = 0.8) +  # Add vertical dashed line at roc_cutoff
    geom_vline(xintercept = log(youden_cutoff), linetype = "longdash", color = "orange", size = 0.8) +  # Add vertical dashed line at roc_cutoff
  labs(
    title = "Put your title here",
    x = "Log Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10), 
    axis.text = element_text(size = 10),  
    plot.title = element_text(size = 10) 
  )
faceted_log_scale
```  

  b. How similar / different is seroprevalance using the different cutoff methods? 
    Edit the code to calculate each of the cutoff methods (some of these you've already calculated in previous questions)

```{r all_seroprevalence}

roc_cutoff <- 550 ##put the cutoff you chose here
seropositivity <- ifelse(sample_data$ZIKA > roc_cutoff, 1, 0) 
 
#number of seropositive people 
table(seropositivity,useNA="always")  
  
#percent of people seropositive and negative
seropos_table <- round(prop.table(table(seropositivity,useNA="always")),3)*100
seropos_table[2]
cat("Seroprevalence:", round(seropos_table[2],4), "%")
  
```

5. Which method do you think you will use going forward in your study (if you will use a cutoff)? Why? 

6. Optional: If you want to apply a mixture model, answer the following questions

  a. Plot your sample values and consider if they have a good distribution to fit a mixture model.

```{r sample_histogram}

  faceted_natural_scale <- ggplot(sample_long, aes(x = mfi)) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  facet_wrap(~ antigen, scales = "free", ncol = 5) +  
  labs(
    title = "",
    x = "Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_natural_scale

#log scale
faceted_log_scale <- ggplot(sample_long, aes(x = log(mfi))) +
  geom_histogram(bins = 30, color = "black", fill = "blue") +
  facet_wrap(~ antigen, scales = "free", ncol = 5) +  # <- Set 5 columns per row
  labs(
    title = "",
    x = "Log Antigen MFI",
    y = "Number of samples"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10),  # Smaller font for facet labels
    axis.text = element_text(size = 10),  # Smaller font for axis text
    plot.title = element_text(size = 10)  # Smaller font for the plot title
  )
faceted_log_scale

```

  b. Decide which antigen you want to test. Make a vector of your antigen values. If you want to do a log transformation, do so before putting data into the FMM. Use the following code to run the model with 2 components
  
```{r fit_fmm_model}
  mfi_values <- sample_data$JE3 #fill in your own data here
  log_antigen_vector <- log(mfi_values)
  k <- 2 #number of components 
  fmm_model <- Mclust(log_antigen_vector, G = k, modelNames = "V") 
  
```
  
  c. Make a plot of your data with the 2 predicted components mapped over them.
  
```{r plot_fmm_components}
    
  # Extract parameters
  means <- fmm_model$parameters$mean
  sds <- sqrt(fmm_model$parameters$variance$sigmasq)
  props <- fmm_model$parameters$pro
  
  # Create density curves for each component
  x_vals <- seq(min(log_antigen_vector), max(log_antigen_vector), length.out = 1000)

  dens_df <- data.frame(
  x = rep(x_vals, 2),
  density = c(
    dnorm(x_vals, mean = means[1], sd = sds[1]) * props[1],
    dnorm(x_vals, mean = means[2], sd = sds[2]) * props[2]
  ),
  component = factor(rep(1:2, each = length(x_vals)))
)

# Plot histogram + density curves
ggplot() +
  geom_histogram(aes(x = log_antigen_vector, y = after_stat(density)), 
                 bins = 50, fill = "gray80", color = "white") +
  geom_line(data = dens_df, aes(x = x, y = density, color = component), size = 1.2) +
  labs(title = "FMM: Mixture of 2 Components",
       x = "log(MFI)", y = "Density") +
  scale_color_manual(values = c("blue", "red")) +
  theme_minimal()
```
  
  d. Test to see what number of components is the best fit for your data. Remember, FMM might not be an appropriate approach if 2 components is not the best fit for your data.
```{r number_components}

# Step 1: Fit BIC values for G = 1 to 4
bic_values <- mclustBIC(log_antigen_vector, modelNames = "V", G = 1:4)

# Step 2: Extract BIC values into a data frame
bic_df <- as.data.frame(bic_values)
bic_df$G <- as.numeric(rownames(bic_df))

# Step 3: Find best model based on highest BIC
best_index <- which.max(bic_df$x)  # V is the column for model "V"

cat("Best number of components based on BIC:", best_index, "\n")

```

  e. If in part d, the model indicates that 2 components is not the best fit for your model, remake the plot showing your sample data fitted with the correct number of components. Copy the code from parts b and c below and edit to run with the best number of components.
  
  f. Based on 2 components, calculate seroprevalence using the code below. Remember, if 2 components is not the best fit for your data, then this estimate may not be valid
  
```{r}

#this uses the fmm_model results you generated in b. 

# Step 1: Out of the 2 components, we identify which distribution has a higher mean, and assume that is the seropositive component
component_means <- fmm_model$parameters$mean
seropositive_component <- which.max(component_means)

# Step 4: Get the posterior probabilities for each sample
posteriors <- fmm_model$z  #For each sample, this is the probability of the component being in the first component or second component
seropositive_probs <- posteriors[, seropositive_component] #this pulls out just the probability of each sample being seropositive

# Step 5: Estimate seroprevalence
estimated_seroprevalence <- mean(seropositive_probs) #the mean of all the probabilities will be equal to the overall population seroprevalence

# Step 6: Report as percentage
cat("Estimated seroprevalence:", round(estimated_seroprevalence * 100, 1), "%\n")
```
  
  g. Compare the seroprevalence you calculated in part f. to the seroprevalence calculated from the other cutoff methods earlier in the lab. 
